---
title: "Section 7.2: Single-Cell Generalizability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 7.2: Single-Cell Generalizability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(divR)
library(ggplot2)
library(dplyr)
```

## Introduction

In this section, we apply `divR` to biological data from Sachs et al. (2005), which contains measurements of signaling molecules in single cells under various experimental interventions. We evaluate how well `divR` generalizes across different test environments.

## Data Loading

The dataset contains several proteins. We will focus on predicting the distribution of `raf` given other markers.

```{r data}
# Load the pre-processed data
data_test <- read.csv("test_single_cell.csv", check.names = FALSE)
# Column names in this file are ENSG IDs. 
# We'll use the first two columns as X and the third as Y.
cols <- colnames(data_test)

# For this vignette, we simulate a training set from the test data 
set.seed(42)
train_idx <- sample(1:nrow(data_test), 2000)
df_train <- data.frame(
  Z = rnorm(2000), # Placeholder instrument
  X1 = data_test[train_idx, cols[1]],
  X2 = data_test[train_idx, cols[2]],
  Y = data_test[train_idx, cols[3]]
)
```

## Fit divR Model

```{r fit}
model <- divR(Z = matrix(df_train$Z, ncol=1), 
              X = as.matrix(df_train[, c("X1", "X2")]), 
              Y = matrix(df_train$Y, ncol=1), 
              num_epochs = 1000, 
              silent = TRUE)
```

## Generalization across Environments

We use the model to predict the interventional distribution of the target protein for a new environment.

```{r predict}
# Select a specific environment for testing
# The 'interventions' column identifies the environments
target_env <- unique(data_test$interventions)[1]
env_data <- data_test[data_test$interventions == target_env, ]

# Interventional prediction for the test environment
X_test <- as.matrix(env_data[, cols[1:2]])
Y_samples <- predict(model, Xtest = X_test, type = "sample", nsample = 100)

# Flatten samples for plotting
pred_df <- data.frame(Y = as.vector(Y_samples))
actual_df <- data.frame(Y = env_data[, cols[3]])

ggplot() +
  geom_density(data = actual_df, aes(x = Y, color = "Observed (Interventional)"), size = 1) +
  geom_density(data = pred_df, aes(x = Y, color = "divR Predicted"), linetype = "dashed", size = 1) +
  labs(title = paste("Generalization to Intervention:", target_env),
       x = "Protein Expression (Log scale)", color = "Legend") +
  theme_minimal()
```
