---
title: "Section 7.1: Colonial Origins of Comparative Development"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 7.1: Colonial Origins of Comparative Development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(divR)
library(ggplot2)
library(dplyr)
library(httr)
library(readstata13)
```

## Introduction

This vignette replicates the analysis of Acemoglu et al. (2001) using the `divR` package. The study investigates the effect of institutional quality (average protection against expropriation risk) on economic performance (log GDP per capita in 1995), using log European settler mortality as an instrument.

## Data Loading

```{r data}
# Load data from the provided URL
url <- "https://www.dropbox.com/scl/fi/3yuv9j514zuajzjfluoc1/maketable4.zip?rlkey=pq9l7bxktw1iqxe6fmoh26g79&e=1&dl=1"
temp_file <- tempfile(fileext = ".zip")
GET(url, write_disk(temp_file, overwrite = TRUE))

temp_dir <- tempdir()
unzip(temp_file, exdir = temp_dir)
dta_file <- list.files(temp_dir, pattern = "\\.dta$", full.names = TRUE)

df <- read.dta13(dta_file)
df <- df %>% filter(baseco == 1)

# Variables
X <- df$avexpr  # Institutional Quality
Z <- df$logem4 # Log Settler Mortality
Y <- df$logpgp95 # Log GDP
```

## Fit divR Model

```{r fit}
# We use a smaller number of epochs for the vignette to speed up rendering
model <- divR(Z = matrix(Z, ncol=1), 
              X = matrix(X, ncol=1), 
              Y = matrix(Y, ncol=1), 
              num_epochs = 5000, 
              silent = TRUE)
```

## Comparison with OLS and 2SLS

```{r plot}
avexpr_seq <- matrix(seq(min(X), max(X), length.out = 100), ncol = 1)

# divR predictions
mean_pred_div <- predict(model, Xtest = avexpr_seq, type = "mean", nsample = 1000)
q_pred_div <- predict(model, Xtest = avexpr_seq, type = "quantile", 
                      quantiles = c(0.1, 0.9), nsample = 1000)

# OLS for comparison
ols_mod <- lm(Y ~ X)
ols_pred <- predict(ols_mod, newdata = data.frame(X = as.vector(avexpr_seq)))

plot_df <- data.frame(
  avexpr = as.vector(avexpr_seq),
  DIV_mean = as.vector(mean_pred_div),
  DIV_lower = q_pred_div[, 1],
  DIV_upper = q_pred_div[, 2],
  OLS = ols_pred
)

ggplot() +
  geom_point(data = df, aes(x = avexpr, y = logpgp95), alpha = 0.5) +
  geom_line(data = plot_df, aes(x = avexpr, y = DIV_mean, color = "divR Mean"), size = 1.2) +
  geom_ribbon(data = plot_df, aes(x = avexpr, ymin = DIV_lower, ymax = DIV_upper), 
              fill = "green", alpha = 0.2) +
  geom_line(data = plot_df, aes(x = avexpr, y = OLS, color = "OLS"), linetype = "dashed") +
  labs(title = "Colonial Origins: divR vs OLS",
       x = "Institutional Quality (avexpr)",
       y = "Log GDP per capita",
       color = "Method") +
  theme_minimal()
```
