---
title: "Section 6.3: Multivariate Endogenous Variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 6.3: Multivariate Endogenous Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(divR)
library(ggplot2)
```

## Introduction

In this section, we demonstrate `divR` on a multivariate treatment case where $X \in \mathbb{R}^2$:
- $X_1$ and $X_2$ are both endogenous (affected by the same hidden confounder $H$).
- The outcome $Y$ depends on both $X_1$ and $X_2$ directly.

## Data Generating Process

- $X_1 = \text{log}(2Z + H + \epsilon_{X1} + 7)$
- $X_2 = Z \cdot (2H - 0.5\epsilon_{X2})$
- $Y = X_1 + 2X_2 + 2H + \epsilon_Y$

```{r dgp}
g_log <- function(Z, H, eps_X) log(2 * Z + H + eps_X + 7)
g_mult <- function(Z, H, eps_X) Z * (2 * H - 0.5 * eps_X)
f_lin <- function(X1, X2, H, eps_Y) 1*X1 + 2*X2 + 2*H + eps_Y

set.seed(1998)
n <- 2000
H <- rnorm(n)
Z <- runif(n, 0, 3)
eps_X1 <- rnorm(n)
eps_X2 <- rnorm(n)
eps_Y <- rnorm(n)

X1 <- g_log(Z, H, eps_X1)
X2 <- g_mult(Z, H, eps_X2)
Y <- f_lin(X1, X2, H, eps_Y)

X <- cbind(X1, X2)
```

## Fit divR Model

```{r fit}
model <- divR(Z = matrix(Z, ncol=1), 
              X = X, 
              Y = matrix(Y, ncol=1), 
              num_epochs = 2000, 
              silent = TRUE)
```

## Interventional Prediction: E[Y | do(X1, X2)]

We vary $X_1$ while keeping $X_2$ fixed at its median to see the partial causal effect.

```{r predict}
x2_fixed <- median(X2)
x1_grid <- seq(min(X1), max(X1), length.out = 100)
X_test <- cbind(x1_grid, rep(x2_fixed, 100))

Y_causal_pred <- predict(model, Xtest = X_test, type = "mean", nsample = 500)

# True mean via large-scale simulation
get_true_mean <- function(x1, x2, n_rep = 5000) {
  H_rep <- rnorm(n_rep)
  eps_Y_rep <- rnorm(n_rep)
  Y_rep <- f_lin(rep(x1, n_rep), rep(x2, n_rep), H_rep, eps_Y_rep)
  mean(Y_rep)
}

true_means <- sapply(x1_grid, function(x1) get_true_mean(x1, x2_fixed))

plot_df <- data.frame(
  X1 = x1_grid,
  Predicted = as.vector(Y_causal_pred),
  True = true_means
)

ggplot(plot_df, aes(x = X1)) +
  geom_line(aes(y = Predicted, color = "divR Estimate"), size = 1) +
  geom_line(aes(y = True, color = "True Causal"), linetype = "dashed", size = 1) +
  labs(title = paste("Partial Effect of X1 (X2 fixed at", round(x2_fixed, 2), ")"),
       y = "E[Y | do(X1, X2)]") +
  theme_minimal()
```
