---
title: "Section 3.3: Replication of Softplus Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 3.3: Replication of Softplus Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(divR)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Methodology: Distributional Instrumental Variables (DIV)

The `divR` package implements **Distributional Instrumental Variables (DIV)**, a method designed to estimate the full interventional distribution $P(Y | do(X = x))$ in the presence of unobserved confounding.

Traditional IV methods like 2SLS focus only on the **Average Treatment Effect (ATE)**, which is the mean of this distribution. However, in many applications, we care about the entire distribution---such as its variance, tail behavior, or specific quantiles.

### The Model Structure
DIV uses a generator-based approach with two neural networks:
1.  **Generator $g(Z, \epsilon_X, \epsilon_H)$**: Models the relationship between the instrument $Z$, the hidden confounder $H$ (represented by $\epsilon_H$), and the treatment $X$.
2.  **Generator $f(X, \epsilon_Y, \epsilon_H)$**: Models the relationship between the treatment $X$, the hidden confounder $H$, and the outcome $Y$.

By sharing the noise component $\epsilon_H$ between both generators, DIV explicitly models the "backdoor" path caused by unobserved confounding. During training, it minimizes an **energy loss** that ensures the generated joint distribution of $(X, Y)$ matches the observed data.

## Data Generating Process (Softplus)

In this example, we use a highly nonlinear "Softplus" DGP. The softplus function $\log(1 + \exp(\cdot))$ is a smooth approximation of the ReLU function, making it a challenging test for linear IV methods.

- **Confounder**: $H \sim \mathcal{N}(0, 1)$
- **Instrument**: $Z \sim \text{Uniform}(0, 3)$
- **Treatment**: $X = \log(1 + \exp(Z + H + \epsilon_X))$
- **Outcome**: $Y = 0.5 \cdot \log(1 + \exp(2X + 2H + 4\epsilon_Y))$

The presence of $H$ in both equations creates a strong correlation between $X$ and $Y$ that is *not* causal. DIV must "unlearn" this correlation to find the true effect of $X$ on $Y$.

```{r dgp}
g_softplus <- function(Z, H, eps_X) log(1 + exp(Z + H + eps_X))
f_softplus <- function(X, H, eps_Y) 0.5 * log(1 + exp(2*X + 2*H + 4*eps_Y))

set.seed(42)
n_train <- 5000
H <- rnorm(n_train)
Z <- runif(n_train, 0, 3)
eps_X <- rnorm(n_train)
eps_Y <- rnorm(n_train)

X <- g_softplus(Z, H, eps_X)
Y <- f_softplus(X, H, eps_Y)
```

## Fitting the Model

We train the DIV model for 2,000 epochs. Because the relationships are nonlinear, we use a sufficiently large number of epochs to allow the neural networks to converge to the correct functional forms.

```{r fit}
model <- divR(Z = matrix(Z, ncol=1),
              X = matrix(X, ncol=1),
              Y = matrix(Y, ncol=1),
              num_epochs = 2000,
              silent = TRUE)
```

## Figure 2: Visualizing the Interventional Distribution

Figure 2 compares the **True Interventional Distribution** $P_Y^{do(X=x)}$ with the **Estimated Interventional Distribution** $\hat{P}_Y^{do(X=x)}$.

- **Points (Blue/Yellow)**: These represent individual samples. For a fixed $X=x$, they show the range of values $Y$ would take if we intervened and set $X$ to that value.
- **Solid Lines (Black/Grey)**: These are the **Causal Quantiles** $q_{\alpha}^*(x)$. We plot the 10th, 50th (median), and 90th percentiles.
- **Goal**: The yellow dots and grey lines (Estimated) should overlap with the blue dots and black lines (True).

```{r figure2}
# 1. Samples for scatter plot
Y_true_int <- f_softplus(X, rnorm(n_train), rnorm(n_train))
Y_est_int <- predict(model, Xtest = matrix(X, ncol=1), type = "sample", nsample = 1)
Y_est_int <- as.vector(Y_est_int)

plot_samples <- data.frame(
  X = rep(X, 2),
  Y = c(Y_true_int, Y_est_int),
  Source = rep(c("True", "Estimated"), each = n_train)
)

X_grid <- matrix(seq(min(X), max(X), length.out = 100), ncol = 1)

# 2. Quantiles for line plot
taus <- c(0.1, 0.5, 0.9)
div_quant <- predict(model, Xtest = X_grid, type = "quantile", quantiles = taus, nsample = 1000)

get_true_quantiles <- function(x_val, n_rep = 5000) {
  H_rep <- rnorm(n_rep); eps_Y_rep <- rnorm(n_rep)
  Y_rep <- f_softplus(rep(x_val, n_rep), H_rep, eps_Y_rep)
  quantile(Y_rep, taus)
}
true_quants <- t(sapply(X_grid, get_true_quantiles))

quant_df <- data.frame(
  X = as.vector(X_grid),
  True_Q10 = true_quants[, 1], True_Q50 = true_quants[, 2], True_Q90 = true_quants[, 3],
  Est_Q10 = div_quant[, 1], Est_Q50 = div_quant[, 2], Est_Q90 = div_quant[, 3]
)

quant_df_long <- quant_df %>%
  pivot_longer(cols = -X, names_to = "Metric", values_to = "Y") %>%
  mutate(
    Source = ifelse(grepl("True", Metric), "Q_True", "Q_Estimated"),
    Quantile = gsub("True_|Est_", "", Metric)
  )

ggplot() +
  geom_point(data = plot_samples, aes(x = X, y = Y, color = Source), alpha = 0.3, size = 0.5) +
  geom_line(data = quant_df_long, aes(x = X, y = Y, group = Metric, color = Source), linewidth = 1) +
  scale_color_manual(
    values = c("True" = "deepskyblue", "Estimated" = "darkgoldenrod1", "Q_True" = "black", "Q_Estimated" = "grey50"),
    labels = c(
      "True" = expression(paste("Sample from ", P[Y]^{do(X==x)})),
      "Estimated" = expression(paste("Sample from ", hat(P)[Y]^{do(X==x)})),
      "Q_True" = expression(q[alpha]^"*" * (x)),
      "Q_Estimated" = expression(hat(q)[alpha]^"*" * (x))
    )
  ) +
  labs(title = "Figure 2: Causal Effect and Distributional Overlap",
       subtitle = "Dots show interventional samples; lines show 10th, 50th, 90th quantiles",
       x = "Treatment (X)", y = "Outcome (Y)", color = "") +
  theme_minimal() +
  theme(legend.text.align = 0)
```

## Figure 3: Density Comparison at Specific Intervention Points

While Figure 2 shows the overall trend, Figure 3 looks at the "shape" of the distribution at specific points. We pick the 25th, 50th, and 75th percentiles of the observed treatment $X$ and perform an intervention $do(X = x)$.

- **Interpretation**: Each panel shows the probability density of $Y$ if *everyone* in the population were assigned the same value of $X$.
- **Key Insight**: Notice how the distribution of $Y$ shifts and potentially changes shape as $X$ increases. DIV accurately captures these distributional shifts, not just the change in mean.

```{r figure3}
x_vals <- quantile(X, c(0.25, 0.5, 0.75))
n_samples_density <- 3000

density_data <- list()
for (i in seq_along(x_vals)) {
  xv <- x_vals[i]
  y_true <- f_softplus(rep(xv, n_samples_density), rnorm(n_samples_density), rnorm(n_samples_density))
  y_est <- predict(model, Xtest = matrix(xv, ncol=1), type = "sample", nsample = n_samples_density)
  density_data[[i]] <- data.frame(
    Y = c(y_true, as.vector(y_est)),
    Source = rep(c("True", "Estimated"), each = n_samples_density),
    X_Val = names(x_vals)[i]
  )
}

plot_density_df <- do.call(rbind, density_data)
plot_density_df$X_Val <- factor(plot_density_df$X_Val, levels = c("25%", "50%", "75%"),
                                labels = c("x[Q25]", "x[Q50]", "x[Q75]"))

ggplot(plot_density_df, aes(x = Y, fill = Source)) +
  geom_density(alpha = 0.5) +
  facet_grid(rows = vars(X_Val), labeller = label_parsed) +
  scale_fill_manual(values = c("True" = "deepskyblue", "Estimated" = "darkgoldenrod1")) +
  labs(title = "Figure 3: Interventional Density Estimates",
       subtitle = "Kernel density comparison at specific intervention values of X",
       x = "Outcome (Y)", y = expression(P[Y]^{do(X==x)})) +
  theme_minimal() +
  xlim(-2, 12)
```
