---
title: "Section 5.2: Binary Treatment and Quantile Effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 5.2: Binary Treatment and Quantile Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(divR)
library(ggplot2)
```

## Introduction

In Section 5.2 of the paper, we explore the performance of DIV in settings with a binary treatment variable. This is a common scenario in causal inference, where the goal is often to estimate the Quantile Treatment Effect (QTE). DIV's ability to model the full interventional distribution makes it particularly well-suited for this task.

## Data Generating Process

We use a logistic DGP where the treatment $D$ is binary and the outcome $Y$ is a nonlinear function of $D$ and a hidden confounder $H$:
- $Z, H, \epsilon_D \sim \text{Logistic}(0, 1)$
- $D = \mathbb{I}(4Z + 4H > \epsilon_D)$
- $Y = \log(1 + \exp(18 + 8D + 6H))$

The nonlinearity in $Y$ ensures that the treatment effect is not constant across the distribution of the outcome.

```{r dgp}
set.seed(12)
n <- 2000
Z <- rlogis(n)
H <- rlogis(n)
D <- as.numeric(4 * Z + 4 * H > rlogis(n))
Y <- log(1 + exp(18 + 8 * D + 6 * H))

df <- data.frame(Z=Z, D=D, Y=Y)
```

## Fit divR Model

```{r fit}
model <- divR(Z = matrix(Z, ncol=1),
              X = matrix(D, ncol=1),
              Y = matrix(Y, ncol=1),
              num_epochs = 2000,
              silent = TRUE)
```

## Quantile Treatment Effect (QTE)

The QTE is defined as the difference between the $\tau$-quantiles of the interventional distributions:
$$QTE(\tau) = Q_\tau(Y | do(D=1)) - Q_\tau(Y | do(D=0))$$

Standard IV methods like 2SLS only estimate the Average Treatment Effect (ATE), which may miss important heterogeneity in how the treatment affects different parts of the population.

```{r qte}
taus <- seq(0.1, 0.9, by = 0.1)

# Predict quantiles for do(D=0) and do(D=1)
q0 <- predict(model, Xtest = matrix(0, nrow=1), type = "quantile", quantiles = taus, nsample = 2000)
q1 <- predict(model, Xtest = matrix(1, nrow=1), type = "quantile", quantiles = taus, nsample = 2000)

qte_pred <- q1 - q0

# True QTE via large-scale simulation (Oracle)
n_oracle <- 10000
H_oracle <- rlogis(n_oracle)
Y0_oracle <- log(1 + exp(18 + 8*0 + 6*H_oracle))
Y1_oracle <- log(1 + exp(18 + 8*1 + 6*H_oracle))

q0_true <- quantile(Y0_oracle, taus)
q1_true <- quantile(Y1_oracle, taus)
qte_true <- q1_true - q0_true

plot_df <- data.frame(
  Tau = taus,
  Estimated = as.vector(qte_pred),
  True = as.vector(qte_true)
)

ggplot(plot_df, aes(x = Tau)) +
  geom_line(aes(y = Estimated, color = "DIV QTE"), linewidth = 1) +
  geom_point(aes(y = Estimated, color = "DIV QTE")) +
  geom_line(aes(y = True, color = "True QTE"), linetype = "dashed", linewidth = 1) +
  geom_point(aes(y = True, color = "True QTE")) +
  scale_color_manual(values = c("DIV QTE" = "darkgoldenrod1", "True QTE" = "black")) +
  labs(title = "Section 5.2: Binary Treatment QTE",
       subtitle = "Comparison of DIV estimated QTE and true QTE",
       x = "Quantile (Tau)", y = "Quantile Treatment Effect",
       color = "Method") +
  theme_minimal()
```

## Conclusion

As shown in the plot, DIV is able to accurately recover the non-constant Quantile Treatment Effect. This demonstrates that DIV successfully identifies the full interventional distribution $P(Y|do(D=d))$ even when the treatment variable is discrete.
