---
title: "Section 5.4: Instrument Strength and Weak Instruments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 5.4: Instrument Strength and Weak Instruments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(divR)
library(ggplot2)
```

## Introduction

In Section 5.4 of the paper, we investigate the sensitivity of DIV to the strength of the instrumental variable $Z$. A key strength of DIV is that it exploits the **full conditional distribution** $P(X|Z)$, whereas some other methods (like standard control functions or 2SLS) rely primarily on the conditional expectation $E[X|Z]$.

Traditional instrumental variable methods often break down when the instrument is "weak" in the sense that $E[X|Z] \approx 0$ (no mean shift). However, DIV remains stable because it uses a conditional generative approach that can capture signal from **any** distributional change, such as changes in variance or shape.

## Data Generating Process

Following Section 5.4 of the paper:
- $Z \sim \text{Uniform}(-3, 3)$
- $H, \epsilon_X, \epsilon_Y \sim \text{Uniform}(-1, 1)$
- $g(Z, H, \epsilon_X) = Z(\alpha + 2H + \epsilon_X)$
- $f(X, H, \epsilon_Y) = \frac{1}{1 + \exp(-(X + 2H + \epsilon_Y)/3)}$

Note that $E[X|Z] = \alpha Z$. Thus, $\alpha$ directly controls the instrument strength in the traditional sense.

```{r dgp}
g_fct <- function(Z, H, eps_X, alpha) Z * (alpha + 2 * H + eps_X)
f_sigmoid <- function(X, H, eps_Y) 1 / (1 + exp(-(X + 2 * H + eps_Y) / 3))

simulate_data <- function(alpha, n = 5000) {
  Z <- runif(n, -3, 3)
  H <- runif(n, -1, 1)
  eps_X <- runif(n, -1, 1)
  eps_Y <- runif(n, -1, 1)
  X <- g_fct(Z, H, eps_X, alpha)
  Y <- f_sigmoid(X, H, eps_Y)
  list(Z=Z, X=X, Y=Y)
}
```

## Comparison: Strong vs. Weak Instrument

We compare the DIV estimates for two scenarios:
1.  **Strong Instrument ($\alpha = 2$):** A traditional strong instrument where $Z$ significantly shifts the mean of $X$.
2.  **Weak Instrument ($\alpha = 0$):** A traditional "useless" instrument where $Z$ has zero effect on the mean of $X$ ($E[X|Z]=0$), but still scales the variance of $X$.

```{r comparison}
set.seed(42)
# Strong Instrument (alpha = 2)
d_strong <- simulate_data(alpha = 2)
mod_strong <- divR(Z = matrix(d_strong$Z, ncol=1),
                   X = matrix(d_strong$X, ncol=1),
                   Y = matrix(d_strong$Y, ncol=1),
                   num_epochs = 500, silent = TRUE)

# Weak Instrument (alpha = 0)
d_weak <- simulate_data(alpha = 0)
mod_weak <- divR(Z = matrix(d_weak$Z, ncol=1),
                 X = matrix(d_weak$X, ncol=1),
                 Y = matrix(d_weak$Y, ncol=1),
                 num_epochs = 500, silent = TRUE)
```

## Results and Theoretical Consistency

In Table 3 of the paper, the **MSE of DIV is 0.002 for both $\alpha=0$ and $\alpha=1$**. In contrast, traditional methods like 2SLS (CF linear) see their MSE jump from 1.832 to **141.941** at $\alpha=0$.

```{r results}
X_grid <- matrix(seq(-4, 4, length.out = 100), ncol = 1)

# True mean via large-scale simulation
get_true_mean <- function(x_val, n_rep = 5000) {
  H_rep <- runif(n_rep, -1, 1); eps_Y_rep <- runif(n_rep, -1, 1)
  mean(f_sigmoid(rep(x_val, n_rep), H_rep, eps_Y_rep))
}
true_means <- sapply(X_grid, get_true_mean)

pred_strong <- predict(mod_strong, Xtest = X_grid, type = "mean")
pred_weak <- predict(mod_weak, Xtest = X_grid, type = "mean")

plot_df <- data.frame(
  X = as.vector(X_grid),
  Strong = as.vector(pred_strong),
  Weak = as.vector(pred_weak),
  True = true_means
)

ggplot(plot_df, aes(x = X)) +
  geom_line(aes(y = Strong, color = "Strong (alpha=2)"), linewidth = 1) +
  geom_line(aes(y = Weak, color = "Weak (alpha=0)"), linewidth = 1) +
  geom_line(aes(y = True, color = "True Causal"), linetype = "dashed", linewidth = 1) +
  scale_color_manual(values = c("Strong (alpha=2)" = "darkgoldenrod1",
                               "Weak (alpha=0)" = "firebrick3",
                               "True Causal" = "black")) +
  labs(title = "Section 5.4: DIV Robustness to Instrument Strength",
       subtitle = "DIV performs equally well when the mean effect alpha is zero",
       y = "E[Y | do(X)]",
       color = "Model") +
  theme_minimal()
```

### Why does the "Weak" instrument case look so accurate?
It may seem counter-intuitive that the "Weak" line is so close to the true causal function. This is exactly what the paper predicts:
1.  **Full Distribution vs. Mean**: While $\alpha=0$ means $Z$ doesn't shift the average value of $X$, the **variance** of $X$ still depends on $Z^2$. DIV's conditional generator successfully "learns" from this variance shift.
2.  **Numerical Stability**: In the "Strong" case ($\alpha=2$), the instrument introduces much more noise and variation into $X$. This larger range can sometimes be slightly more challenging to optimize for a fixed number of epochs.
3.  **Identifiability**: The paper proves that the interventional distribution is identifiable even when the mean relevance is zero, provided there is enough "richness" in the conditional distribution $P(X|Z)$.

## Conclusion

The simulation study confirms that DIV is **robustly superior** to traditional methods in the face of weak instruments. While methods that rely on $E[X|Z]$ break down completely as $\alpha \to 0$, DIV maintains near-perfect estimation accuracy by leveraging the information contained in the higher-order moments of the treatment distribution.
