---
title: "Section 6.4: Instrument Strength and Weak Instruments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 6.4: Instrument Strength and Weak Instruments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(divR)
library(ggplot2)
```

## Introduction

Instrumental variable estimation relies on the instrument $Z$ being relevant to the treatment $X$. In this section, we explore how `divR` performs when the instrument is weak.

## Data Generating Process

We use a parameter $\alpha$ to control the instrument strength:
- $X = \alpha(Z + H) + 0.5\epsilon_X$
- $Y = \tanh(1.5X) - 2.5H + \epsilon_Y$

When $\alpha$ is small, the instrument is "weak".

```{r dgp}
g_fct <- function(Z, H, eps_X, alpha) alpha * (Z + H) + 0.5 * eps_X
f_tanh <- function(X, H, eps_Y) tanh(1.5 * X) - 2.5 * H + eps_Y

simulate_data <- function(alpha, n = 2000) {
  H <- runif(n, -1, 1)
  Z <- runif(n, -3, 3)
  eps_X <- runif(n, -1, 1)
  eps_Y <- runif(n, -1, 1)
  X <- g_fct(Z, H, eps_X, alpha)
  Y <- f_tanh(X, H, eps_Y)
  list(Z=Z, X=X, Y=Y)
}
```

## Comparison: Strong vs Weak Instrument

 we fit the model for $\alpha = 5$ (strong) and $\alpha = 0.5$ (weak).

```{r comparison}
# Strong Instrument (alpha = 5)
d_strong <- simulate_data(alpha = 5)
mod_strong <- divR(Z = matrix(d_strong$Z, ncol=1), X = matrix(d_strong$X, ncol=1), Y = matrix(d_strong$Y, ncol=1), num_epochs = 1000, silent = TRUE)

# Weak Instrument (alpha = 0.5)
d_weak <- simulate_data(alpha = 0.5)
mod_weak <- divR(Z = matrix(d_weak$Z, ncol=1), X = matrix(d_weak$X, ncol=1), Y = matrix(d_weak$Y, ncol=1), num_epochs = 1000, silent = TRUE)
```

## Results

```{r results}
X_grid <- matrix(seq(-2, 2, length.out = 100), ncol = 1)

# True mean
get_true_mean <- function(x_val, n_rep = 5000) {
  H_rep <- runif(n_rep, -1, 1)
  eps_Y_rep <- runif(n_rep, -1, 1)
  mean(f_tanh(rep(x_val, n_rep), H_rep, eps_Y_rep))
}
true_means <- sapply(X_grid, get_true_mean)

pred_strong <- predict(mod_strong, Xtest = X_grid, type = "mean")
pred_weak <- predict(mod_weak, Xtest = X_grid, type = "mean")

plot_df <- data.frame(
  X = as.vector(X_grid),
  Strong = as.vector(pred_strong),
  Weak = as.vector(pred_weak),
  True = true_means
)

ggplot(plot_df, aes(x = X)) +
  geom_line(aes(y = Strong, color = "Strong Instrument (alpha=5)"), size = 1) +
  geom_line(aes(y = Weak, color = "Weak Instrument (alpha=0.5)"), size = 1) +
  geom_line(aes(y = True, color = "True Causal"), linetype = "dashed", size = 1) +
  labs(title = "Effect of Instrument Strength on divR Estimates",
       y = "E[Y | do(X)]", color = "Model") +
  theme_minimal()
```
